---
title: "NBA Submission File"
author: "Sarah Bailey 301305478"
date: "11/14/2016"
output: 
  html_document:
      toc: yes
      toc_float: yes
      toc_depth: 4
      collapsed: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(fields, plotrix, dplyr,plyr, glmnet,caret,forcats,xgboost, dplyr,aod,mboost,Matrix, e1071,ggplot2)
```
### Introduction
The purpose of this competition was to compute the likelihood that a basketball player makes a shot. This second week, I decided to look at the player characteristics, non-linear effects, and opted for an Xgboost model over the glmnet. In the end, my best score came from using Lucas's code for his data cleaning, however, this was only better than my submission by .00064. I include both here for comparision of models.

```{r eval_metric, include = F}
eval_metric <- function(response, predicted_probs) {
  N <- length(response)
  logloss <- -(1/N) * sum(response * log(predicted_probs) + (1-response) * log(1 - predicted_probs))
  logloss
}

# Function that converts factor to numeric variable
fac_to_num <- function(fac) {
  as.numeric(as.character(fac))
}
```

```{r load_data, include = FALSE}
train <- read.csv("./data/train.csv", stringsAsFactors = F)
test <- read.csv("./data/test.csv", stringsAsFactors = F)
players <- read.csv("./data/players.csv", stringsAsFactors = F)
teams <- read.csv("./data/teams.csv", stringsAsFactors = F)
```


```{r plot_court_function, include = F}
## draw the basketball court 
draw.court = function() {
  rect(0, 0, 94, 50)
  circle = function(x, y, r, from = 0, to = 2 * pi, lines = FALSE, ...) {
    theta = seq(from, to, length = 100)
    if (lines)
      lines(x + r * cos(theta), y + r * sin(theta), ...)
    else polygon(x + r * cos(theta), y + r * sin(theta), ...)
  }
  points(c(5.25, 94 - 5.25), c(25, 25), cex = 2)
  segments(47, 0, 47, 50)
  circle(47, 25, 8)
  circle(47, 25, 2, col = "lightgray")
  theta1 = acos((25 - 35/12)/23.75)
  circle(5.25, 25, 23.75, -pi/2 + theta1, pi/2 - theta1, TRUE)
  circle(94 - 5.25, 25, 23.75, pi/2 + theta1, 3 * pi/2 - theta1, TRUE)
  segments(0, 35/12, 5.25 + 23.75 * sin(theta1), 35/12)
  segments(0, 50 - 35/12, 5.25 + 23.75 * sin(theta1), 50 - 35/12)
  segments(94, 35/12, 94 - 5.25 - 23.75 * sin(theta1), 35/12)
  segments(94, 50 - 35/12, 94 - 5.25 - 23.75 * sin(theta1), 50 - 35/12)
  circle(19, 25, 6, -pi/2, pi/2, TRUE)
  circle(19, 25, 6, pi/2, 3 * pi/2, TRUE, lty = 2)
  circle(94 - 19, 25, 6, pi/2, 3 * pi/2, TRUE)
  circle(94 - 19, 25, 6, -pi/2, pi/2, TRUE, lty = 2)
  circle(5.25, 25, 4, -pi/2, pi/2, TRUE)
  circle(94 - 5.25, 25, 4, pi/2, 3 * pi/2, TRUE)
  rect(0, 17, 19, 33, border = "gray")
  rect(94, 17, 94 - 19, 33, border = "gray")
}

# Function that plots the court
plot_court <- function(main) {
  plot(0,0,pch=46,xlim=c(0,94), ylim=c(0,50), main=main, xlab = '', ylab = '')
  draw.court()
}
```


```{r calc_dist, include = F}
# X,Y coordinates for the basketball hoops
left_hoop_xy <- cbind(5.25, 25)
right_hoop_xy <- cbind(88.75, 25)

# Function to calculate distance between shot location and hoop
calc_shot_dist <- function(shot_data) {
  n <- nrow(shot_data)
  left_shot_ind <- shot_data$offense_basket == "L"
  right_shot_ind <- shot_data$offense_basket == "R"
  shot_dist <- numeric(n)
  shot_dist[left_shot_ind] <- 
    rdist(shot_data[left_shot_ind, c("shot_x", "shot_y")], left_hoop_xy)
  shot_dist[right_shot_ind] <- 
    rdist(shot_data[right_shot_ind, c("shot_x", "shot_y")], right_hoop_xy)
  shot_dist
}
```


```{r data_prep,cache=TRUE}
train$shot_dist <- calc_shot_dist(train)
test$shot_dist <- calc_shot_dist(test)
full <- rbind(train,test)
```

### Revamped 
The following is an edited version of Lucas/Dani
```{r dani_code,echo=T}
#Contested Shot, 0 for not contested, 1 for contested, 2 for Wide Open
full$contested=0

full[which(full$ndd<4),]$contested=1

full[which(full$ndd>10),]$contested=2

#Summarize the shot clock, missing data is becuase shot clock was turned off due to game clock having less than 24 seconds on it
full[is.na(full$shot_shot_clock),]$shot_shot_clock=full[is.na(full$shot_shot_clock),]$shot_game_clock

#if shot was taken in regular flow of offense rushed =0
full$rushed=0

#if shot clock was under 7 seconds set to 1
full[which(full$shot_shot_clock<7),]$rushed=1

#if shot clock was over 19, it was probably a fast break, and a good shot
full[which(full$shot_shot_clock>19),]$rushed=2

#was the shot a catch and shoot
#if the shooter didnt dribble then set to 1
full$catchnshoot=0

full[which(full$dribbles==0),]$catchnshoot=1

#Catch and Shoot
full$ispasser=1

full[is.na(full$passer),]$ispasser=1

##Create Left Shot x and y variables for test and full
full$LeftShotx=NA
full$LeftShoty=NA

full$LeftShotx[full$offense_basket == "L"]=(full$shot_x[which(full$offense_basket == "L")])
full$LeftShoty[full$offense_basket == "L"]=full$shot_y[which(full$offense_basket == "L")]

full$LeftShotx[full$offense_basket == "R"]=94-(full$shot_x[which(full$offense_basket == "R")])
full$LeftShoty[(full$offense_basket == "R")]=50-(full$shot_y[which(full$offense_basket == "R")])


## Change Foul shots that don't have the offense hoop vairable switched
full$LeftShotx[which(full$LeftShotx>73 & full$LeftShotx<75 & full$LeftShoty>23 & full$LeftShoty<27)]=19.75
full$LeftShoty[which(full$LeftShotx>73 & full$LeftShotx<75 & full$LeftShoty>23 & full$LeftShoty<27)]=25

## Label as LayUp, BaseMid, ElbowMid, Long2, Corner3, Break3, Other, weirdxy, Foul Shots
full$Location=NA


full[which(full$LeftShotx>0 & full$LeftShotx<14 & full$LeftShoty<20 & full$shot_dist<19),]$Location="BaseMid"
full[which(full$LeftShotx>0 & full$LeftShotx<14 & full$LeftShoty<47 & full$shot_dist<19 ),]$Location="BaseMid"

full[which(full$shot_dist>8 & full$shot_dist<19 &full$LeftShotx>14) ,]$Location="ElbowMid"

full[which(full$LeftShotx>0 & full$LeftShotx<14 & full$shot_dist>19 &full$shot_dist<22),]$Location="Long2"
full[which(full$shot_dist>19 &full$shot_dist<23.75 &full$LeftShotx>14) ,]$Location="Long2"

full[which(full$LeftShotx>0 & full$LeftShotx<14 & full$LeftShoty<3),]$Location="Corner3"
full[which(full$LeftShotx>0 & full$LeftShotx<14 & full$LeftShoty>47 ),]$Location="Corner3"
full[which(full$LeftShotx>0 & full$LeftShotx<14 & full$shot_dist>22 ),]$Location="Corner3"

full[which(full$shot_dist<28 & full$shot_dist>23.75 &full$LeftShotx>14) ,]$Location="Break3"

full[which(full$shot_dist>28 ) ,]$Location="Other"

full[is.na(full$Location),]$Location="Weirdxy"

full[which(full$LeftShotx<21 & full$LeftShotx>19 & full$LeftShoty>23 & full$LeftShoty<27),]$Location="FreeThrow"

full[which(full$LeftShotx>0 & full$LeftShotx<14 & full$LeftShoty>20 & full$LeftShoty<30),]$Location="LayUp"
```


```{r data, echo=T}
# changed the name of data
changed_name <- function(prefix,data){
  new_names <- rep(NA,ncol(data))
  for (i in 1:length(names(data))){
    new_names <- paste(prefix,names(data)[i],sep="_")
    names(data)[i] <- new_names
  }
  return(data)
}

# Merged variables from player file
shooter_merged <- players
shooter_merged$X <- NULL
shooter_merged <- changed_name("SH",shooter_merged)

passer_merged <- players
passer_merged$X <- NULL
passer_merged <- changed_name("PS",passer_merged)

defender_merged <- players
defender_merged$X <- NULL
defender_merged <- changed_name("DEF",defender_merged)

### position
full <- left_join(full,shooter_merged,by=c("shooter" = "SH_player_id"))
full <- left_join(full,passer_merged,by=c("passer" = "PS_player_id"))
full <- left_join(full,defender_merged,by=c("defender" = "DEF_player_id"))

full$shooter_age <- full$season-full$SH_byear
full$shooter_exp <- full$season-full$SH_rookie

### free throw
FT <- subset(full,(shot_shot_clock ==24 & (shot_dist <=17.5 & shot_dist >=13))|
               (shot_shot_clock ==24 & (shot_dist <69 & shot_dist >=67.5)))

full$free_throw <- ifelse(full$id %in% FT$id,1,0)

full$dribbles <- ifelse(full$free_throw==1,1,full$dribbles)

### shot clock effect
full$shot_shot_clock[(full$shot_shot_clock)<0] <- 0
full$shot_shot_clock[(full$shot_shot_clock)>24] <- 24
full$shot_shot_clock[(is.na(full$shot_shot_clock))] <- mean(full$shot_shot_clock[full$shot_shot_clock<24],na.rm=T)
full$shot_clock_group <- cut(full$shot_shot_clock , c(0,4,7,15,18,22,25), right = FALSE,dig.lab = 4)

### team effect
full$team_vs_opp <- ifelse(full$team>full$opponent,paste(full$team,"vs",full$opponent),paste(full$opponent,"vs",full$team))

### dribble effect
full$dribble_group <- cut(full$dribbles , c(0,1,2,3,6,36), right = FALSE,dig.lab = 4)

full$shot_dist_group <- cut(full$shot_dist , c(0,6,10,15,19,23.74,32,100), right = FALSE,dig.lab = 4)

full[is.na(full$pass_distance),]$pass_distance <- 0

# change the NA to missing
full$SH_position <- fct_explicit_na(full$SH_position,"missing")

# replace NA with its mean
NAlist <- c("ndd","SH_height","SH_weight","SH_byear","SH_rookie","shooter_age","shooter_exp")
for(i in 1:length(NAlist)){
  full[is.na(full[,NAlist[i]]), NAlist[i]] <- mean(full[,NAlist[i]], na.rm = TRUE)
}

```

```{r matrix_prep,echo=T}
train.exp <- full[1:nrow(train),]
test.exp <- full[(nrow(train)+1):nrow(full),]

full.m <-  full %>% dplyr::select((season),(quarter),(home),pass_distance,(shooter),shot_shot_clock,dribbles,distance_travelled,ndd,shot_dist,contested,rushed,catchnshoot,Location,SH_position,SH_height,SH_weight,shooter_age,shooter_exp,free_throw,shot_clock_group,team_vs_opp,dribble_group,shot_dist_group)

# Sparse Matrix
full.sparse <- sparse.model.matrix(~factor(season)+factor(quarter)+factor(team_vs_opp)+factor(home)+pass_distance+factor(shooter)+shot_shot_clock+dribbles+distance_travelled+ndd+shot_dist+factor(contested)+factor(rushed)+factor(catchnshoot)+factor(Location)+factor(SH_position)+SH_height+SH_weight+shooter_age+shooter_exp+factor(free_throw)+factor(shot_clock_group)+factor(team_vs_opp)+factor(dribble_group)+factor(shot_dist_group)-1,data=full.m)



X <- full.sparse[1:nrow(train),]
X_test <- full.sparse[(nrow(train)+1):nrow(full),]

y <- train$made

```

```{r xgboost_2, include=F}
## Xgboost model from Sunghu Code
system.time(
  {
    bst <- xgboost(data = X,label=y, eval_metric ="logloss",
                   max.depth = 4, eta = .1, nround = 750,gamma=1, #750 round takes just about 5 mins
                   colsample_bytree = 1,min_child_weight = 95, # 95 is better than 45, 45 is better than 295
                   nthread = 2, objective = "binary:logistic")
    pred <- predict(bst, X_test)
  }
)

submission_matrix <- data.frame(test.exp$id, pred)
names(submission_matrix) = c('id', 'made')

# Write submission file
write.csv(submission_matrix, file='submission_file3.csv', row.names = FALSE)
```

```{r xgboost, eval=F}
## Xgboost model from Sunghu Code
system.time(
  {
    bst <- xgboost(data = X,label=y, eval_metric ="logloss",
                   max.depth = 4, eta = .1, nround = 750,gamma=1, #750 round takes just about 5 mins
                   colsample_bytree = 1,min_child_weight = 95, # 95 is better than 45, 45 is better than 295
                   nthread = 2, objective = "binary:logistic")
    pred <- predict(bst, X_test)
  }
)

submission_matrix <- data.frame(test.exp$id, pred)
names(submission_matrix) = c('id', 'made')

# Write submission file
write.csv(submission_matrix, file='submission_file3.csv', row.names = FALSE)
```

### Data Exploration and Analysis 
The following is the second best entry, that I had more spent most of my time working on. The initial set up was the same, but I had added different variables and took non linear effects into consideration.

```{r second_best, include=F}
full2 <- rbind(train,test)

library(dplyr)
full2 <- full2 %>% left_join(players, by = c("shooter" = "player_id"))
full2 <- full2 %>% left_join(teams, by = c("team" = "global_id"))
full2 <- full2 %>% left_join(teams, by = c("opponent" = "global_id"))

full2$LeftShotx=NA
full2$LeftShoty=NA

full2$LeftShotx[full2$offense_basket == "L"]=(full2$shot_x[which(full2$offense_basket == "L")])
full2$LeftShoty[full2$offense_basket == "L"]=full2$shot_y[which(full2$offense_basket == "L")]

full2$LeftShotx[full2$offense_basket == "R"]=94-(full2$shot_x[which(full2$offense_basket == "R")])
full2$LeftShoty[(full2$offense_basket == "R")]=50-(full2$shot_y[which(full2$offense_basket == "R")])


## Change Foul shots that don't have the offense hoop vairable switched
full2$LeftShotx[which(full2$LeftShotx>73 & full2$LeftShotx<75 & full2$LeftShoty>23 & full2$LeftShoty<27)]=19.75
full2$LeftShoty[which(full2$LeftShotx>73 & full2$LeftShotx<75 & full2$LeftShoty>23 & full2$LeftShoty<27)]=25

## Label as LayUp, BaseMid, ElbowMid, Long2, Corner3, Break3, Other, weirdxy, Foul Shots
full2$Location=NA


full2[which(full2$LeftShotx>0 & full2$LeftShotx<14 & full2$LeftShoty<20 & full2$shot_dist<19),]$Location="BaseMid"
full2[which(full2$LeftShotx>0 & full2$LeftShotx<14 & full2$LeftShoty<47 & full2$shot_dist<19 ),]$Location="BaseMid"

full2[which(full2$shot_dist>8 & full2$shot_dist<19 &full2$LeftShotx>14) ,]$Location="ElbowMid"

full2[which(full2$LeftShotx>0 & full2$LeftShotx<14 & full2$shot_dist>19 &full2$shot_dist<22),]$Location="Long2"
full2[which(full2$shot_dist>19 &full2$shot_dist<23.75 &full2$LeftShotx>14) ,]$Location="Long2"

full2[which(full2$LeftShotx>0 & full2$LeftShotx<14 & full2$LeftShoty<3),]$Location="Corner3"
full2[which(full2$LeftShotx>0 & full2$LeftShotx<14 & full2$LeftShoty>47 ),]$Location="Corner3"
full2[which(full2$LeftShotx>0 & full2$LeftShotx<14 & full2$shot_dist>22 ),]$Location="Corner3"

full2[which(full2$shot_dist<28 & full2$shot_dist>23.75 &full2$LeftShotx>14) ,]$Location="Break3"

full2[which(full2$shot_dist>28 ) ,]$Location="Other"

full2[is.na(full2$Location),]$Location="Weirdxy"

full2[which(full2$LeftShotx<21 & full2$LeftShotx>19 & full2$LeftShoty>23 & full2$LeftShoty<27),]$Location="FreeThrow"

full2[which(full2$LeftShotx>0 & full2$LeftShotx<14 & full2$LeftShoty>20 & full2$LeftShoty<30),]$Location="LayUp"

full2$Location<-as.factor(full2$Location)

FT2 <- subset(full2,(shot_shot_clock ==24 & (shot_dist <=17.5 & shot_dist >=13))|
               (shot_shot_clock ==24 & (shot_dist <69 & shot_dist >=67.5)))

full2$free_throw <- ifelse(full2$id %in% FT2$id,1,0)
```

```{r Non_linear, include=T}
#Dealing wiht nonlinear variables
full2$shot_shot_clock<-ifelse(is.na(full2$shot_shot_clock),100, full2$shot_shot_clock)
full2$dummy<-factor(full2$shot_shot_clock==100)
full2$SC2<-I(full2$shot_shot_clock^2)
full2$SC3<-I(full2$shot_shot_clock^3)
full2$SDLong<-I(full2$shot_dist>25)
full2$ndd<-ifelse(full2$ndd>15 | is.na(full2$ndd), 0, full2$ndd)
full2$ndd2<-I(full2$ndd^2)
full2$ndd3<-I(full2$ndd^3)

#Binary variable to determine if player is new or not (<=3 years considered "rookie")
is.rookie <- function(player_data) {
x<-player_data$season-player_data$rookie
x2<-ifelse(x<=3, 1, 0)
x2
}
full2$newplayer<-is.rookie(full2)
full2$newplayer<-ifelse(is.na(full2$newplayer),0,full2$newplayer)

# Creating New test and train data
train.exp2 <- full2[1:nrow(train),]
test.exp2 <- full2[(nrow(train)+1):nrow(full2),]
```

### Visualizations
Here are some graphs that give some insight into the types of variables we have and the differences between shot success for different factor level variables

```{r Visual, include=T}
#Visualizing non linear data for ndd variable
train.exp2$ndd2 <- round(train.exp2$ndd,0)
explore.ndd <- train.exp2 %>% dplyr::group_by(ndd2)%>% 
  dplyr::summarise(FG_Per = mean(made))
plot(explore.ndd[1:20,],type="l",main="FG% by ndd",xlab="ndd",ylab="FG%")
lines(predict(loess(explore.ndd$FG_Per~explore.ndd$ndd2)), col='red', lwd=2)
legend(x=17, y=0.38, legend=c("Original", "Smoothing"), 
       lty="solid", col=c("black","red"))

#Create a histogram that models player position, and shot distance (only consider players with >=300 shots)
train.exp1<-subset(train.exp2, count(shooter)>=300)
ggplot(train.exp1, aes(x = as.factor(made), y = shot_dist, fill = as.factor(made))) + geom_boxplot() +
  facet_wrap(~ position, ncol = 5)
```

We can see that different positions have more distance range with shot success, conversely, different positions also have less success with different shot distances.

### Building the Model Part Two

```{r NA, include=T}
#Getting rid of NA's in testing and training data
train.exp2$dribbles<-ifelse(is.na(train.exp2$dribbles),mean(train.exp2$dribbles,na.rm=T),train.exp2$dribbles)
train.exp2$ndd<-ifelse(is.na(train.exp2$ndd), mean(train.exp2$ndd,na.rm=T),train.exp2$ndd)
train.exp2$pass_distance<-ifelse(is.na(train.exp2$pass_distance), mean(train.exp2$pass_distance, na.rm=T), train.exp2$pass_distance)
train.exp2$passer<-ifelse(is.na(train.exp2$passer), 0, train.exp2$passer)
train.exp2$position<-ifelse(is.na(train.exp2$position),"none", train.exp2$position)

test.exp2$dribbles<-ifelse(is.na(test.exp2$dribbles),mean(test.exp2$dribbles,na.rm=T),test.exp2$dribbles)
test.exp2$ndd<-ifelse(is.na(test.exp2$ndd), mean(test.exp2$ndd,na.rm=T),test.exp2$ndd)
test.exp2$pass_distance<-ifelse(is.na(test.exp2$pass_distance), mean(test.exp2$pass_distance, na.rm=T), test.exp2$pass_distance)
test.exp2$passer<-ifelse(is.na(test.exp2$passer),0,test.exp2$passer)
test.exp2$position<-ifelse(is.na(test.exp2$position),"none", test.exp2$position)
```

```{r sparse, include=T}

# The following formula was used in the second best submission
y <- train.exp2$made
X2<-sparse.model.matrix(~shot_dist+ team+opponent+shooter+ndd+ndd2+ndd3+shot_dist+position+ dribbles+
                          Location+shot_shot_clock+SC2+SC3+SDLong:position+free_throw,data=train.exp2)

```